Bulk Semantic Annotation with a Partially Known Knowledge Base\ednote{ Too \snify specific... }

### 2025-01-06 doc/cicm25-sysdesc/sysdesc.tex
paper

\newcommand\ALeA{\textsf{ALeA}\xspace}
\newcommand\snify{\textsf{snify}\xspace}
\def\llangle{\langle\kern-.2em\langle}
\def\rrangle{\rangle\kern-.2em\rangle}

\title{Bulk Semantic Annotation with a Partially Known Knowledge Base}
\author{Michael Kohlhase, Jan Frederik Schaefer}
\institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
\begin{document}
\maketitle
\begin{abstract}
  Semantic Annotations can be utilized for learning support services in technical
  documents. The resulting active documents support more direct and personalized
  interaction with the knowledge conveyed, but require a large investment in the authoring
  process.

@@

\section{Introduction}
Arguably, dealing with large document collections is one of the key factors in the
knowledge-driven society and economy. There are currently two main contenders for machine
support in this area. Symbolic/logic-based technologies and sub-symbolic,
machine-learning-based AI, e.g.\ via LLMs or chatbots. They have complementary strengths
and challenges: Symbolic technologies offer precision and explainability out of the box,
but face scalability challenges because the prerequisite background knowledge has to be
(manually) formalized. ML-based approaches can be trained on all data of the Internet, but
face challenges in precision and explainability. In \cite{Ranta:atcp17},
Aarne Ranta matches these profiles to the notion of \textbf{producer tasks} -- i.e.\ tasks
where precision is key, but limited coverage ($10^{3\pm1}$ concepts) is OK;
e.g.\ for multi-language/variant manuals for very
expensive machines -- and \textbf{consumer tasks}, where coverage is key and
precision secondary; e.g.\ for consumer-grade machine translation like \textsf{Google translate}.

In this paper, we address tools that help attack the coverage problem in symbolic/semantic
approaches in practice. The producer task we use as a case study is that of adaptive
learning assistants for STEM subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
\ALeA uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
variant of {\LaTeX} that allows to embed semantic annotations -- for knowledge
representation, and generates learner-adaptive learning objects instrumented with
learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
\ednote{MK@MK: continue to describe the size and distribution, and
  authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
  \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
  relative sizes.}

We contend that while education -- especially higher education -- as a whole is a
wide-coverage task for society education of individual learners in specific domains --
especially, if we want to tailor the educational offerings to said individual/domain -- is
a producer task, at least until ML-based methods reach the precision and explainability
required by the ethics of teaching.

\section{Semantic Authoring in a Content Commons}

@@

\subsection{Running Example: Annotating  ``Terminal''}
The main practical problem of annotating presentational {\LaTeX} course materials into
semantic documents that carry enough information to support meaningful learning support
services is to add semantic references to technical terms from the domain jargon.
% TODO: This the following better?
% With \sTeX, authors can add semantic references to technical terms,
% which enables straight-forward learning support services like
% showing the definition of a term, but also informs the system
% about the content/prerequisites of a learning object.
% Adding these references is a key practical problem when converting presentational {\LaTeX}
% course materials into semantic documents.
Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
the term ``terminal'' with the (semantic) concept of ``a goal state of a search
problem''. The latter can either be defined earlier in the course or in the domain model
(or both).

@@

@@

To annotate the string ``terminal'' in the learning object, the author has to be aware
that it is a technical term, that the module from \cref{fig:state-space} exists, know the
symbol name and its URL\ednote{path/URI/identifier?}, and manage redundancy of the imports -- i.e. only
adding the \lstinline|\importmodule| directive if it is not already (recursively) implied
and possibly removing directives that become redundant by the new one.  At ca 150,000
words in a semester's words of lecture notes and ca. 10--15\%\ednote{can we actually
  measure this for AI?} of technical terms a rather daunting task.

\sTeX authoring is already supported by a VSCode IDE plugin \cite{sTeX-IDE:git} that
analyzes annotations on the fly, displays the underlying reference semantics, reports
errors and redundancies, and even offers a concept search interface.
But these features only support already annotated terms, so they do not solve the annotation problem above.

\subsection{Semantic Authoring as a Distinct Problem}

We contend that semantic authoring is a problem that is distinct from traditional
(informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
while sharing some of the aspects of either.

@@

\paragraph{Semantic Authoring}
In semantic authoring, e.g. the \sTeX sources of educational content for the \ALeA system
presented above, the base of the text is classical narrative text to which most of the
constraints of classical authoring apply. The exception is the authoring-time tailoring to
the competencies of the assumed reader, which can (partially) be automated by systems like
\ALeA. Note that it is not a restriction that \ALeA is initially targeted to tertiary
education; the underlying mechanisms generalize to all scientific communication and
technical documents. After all, the purpose is to transfer information (instruct), even if
the roles of instructor and instructee are not predetermined by academic seniority.

@@

\paragraph{Semantic Authoring in \sTeX}
How this can concretely work depends on the particular semantic framework, the specific
annotations we want to support, and the semantics embedded in the content commons. In the
\sTeX framework, where we semantically annotate formulae and technical terms with pointers
to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
in sentences and glyphs in formulae with the verbalizations and notations in the content
commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
formulae via notations has been presented by Vre{\v{c}}ar et al. in
\cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
annotation of technical terms via verbalizations.

\section{The \snify System}

@@

\section{Practical Evaluation}

@@

\section{Conclusion}

>>>>

\newcommand\ALeA{\textsf{ALeA}\xspace}
\newcommand\SMGloM{\textsf{SMGloM}\xspace}
\newcommand\snify{\textsf{snify}\xspace}
\def\llangle{\langle\kern-.2em\langle}
\def\rrangle{\rangle\kern-.2em\rangle}

\title{Bulk Semantic Annotation with a Partially Known Knowledge Base\ednote{
    Too \snify specific...
}}
\author{Michael Kohlhase, Jan Frederik Schaefer}
\institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
\begin{document}
\maketitle
\begin{abstract}
  Semantic annotations can be utilized for learning support services in technical
  documents. The resulting active documents support more direct and personalized
  interaction with the knowledge conveyed, but require a large investment in the authoring
  process.

@@

\section{Introduction}
Arguably, dealing with large document collections is one of the key factors in the knowledge-driven society and economy.
There are currently two main contenders for machine support in this area.
Symbolic/logic-based technologies and sub-symbolic/machine-learning-based AI, e.g.\ via LLMs or chatbots.
They have complementary strengths and challenges:
ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
Symbolic technologies, on the other hand, offer precision and explainability out of the box, but face scalability challenges
-- in particular because of the cost of semantic authoring\ednote{
    \emph{semantic authoring} already used in context of semantic web (RDFa).
    Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
    We have a module system and need more dependency management, something else?
    This has to be described somewhere.

    Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
}, which we discuss in this paper.

As a case study, we will use adaptive learning assistants for STEM subjects in higher education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
We contend that machine-learning based technologies are largely unsuitable for this task -- at least until they reach the
precision and explainability required by the ethics of teaching --
while the scope of specific domains is just narrow enough that symbolic approaches are viable.
Conventional (\LaTeX-based) lecture materials can essentially be directly imported into \ALeA,
but its ability to generate learner-adaptive learning objects -- instrumented with learning support interactions --
depends on semantic annotations.
These can be embedded \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
In this paper, we will mostly focus on annotating terms with semantic concept they refer to,
but \sTeX supports a much wider range of annotations, such as annotating problems with the learning objectives
or annotating the defining phrase (definiens) in concept definitions,
and can, in principle, even be used for full formalization.\ednote{True?}
As \sTeX supports the full spectrum from informal (unannotated) to fully formalized content,
we also call \sTeX content \textbf{flexiformal}.

The \textbf{\sTeX/\ALeA content commons} consists of \SMGloM~\cite{GinIanJuc:spsttom16},
the domain model for semantic annotations,
as well as re-usable lecture materials such as lecture slides and problems -- mostly from the area of computer science.
The sheer size of the commons is one of the key challenges for authors:
\SMGloM contains definitions of $X$\ednote{TODO} STEM concepts in multiple languages,
created and maintained by over 30 contributors.
Additionally, there are roughly 4900\ednote{
    Includes 440 from UPS, which seems unannotated... Other issues?
    Command (tex comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
}
semantically annotated lecture slides and 4000 problems.

\paragraph{Overview}
We argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
but is nevertheless a distinct problem requiring special tool support (\cref{sec:semauth}).
Afterwards, we discuss (potential) tools in various stages of development (\cref{sec:tools}) and then
describe and evaluate one concrete tool, \snify, that supports the annotation
of terms with the semantic concepts they refer to (\cref{sec:snify,sec:eval}).
The paper ends with a brief conclusion (\cref{sec:conclusion}).

% % OLD INTRO
% Arguably, dealing with large document collections is one of the key factors in the
% knowledge-driven society and economy. There are currently two main contenders for machine
% support in this area. Symbolic/logic-based technologies and sub-symbolic,
% machine-learning-based AI, e.g.\ via LLMs or chatbots. They have complementary strengths
% and challenges: Symbolic technologies offer precision and explainability out of the box,
% but face scalability challenges because the prerequisite background knowledge has to be
% (manually) formalized. ML-based approaches can be trained on all data of the Internet, but
% face challenges in precision and explainability. In \cite{Ranta:atcp17},
% Aarne Ranta matches these profiles to the notion of \textbf{producer tasks} -- i.e.\ tasks
% where precision is key, but limited coverage ($10^{3\pm1}$ concepts) is OK;
% e.g.\ for multi-language/variant manuals for very
% expensive machines -- and \textbf{consumer tasks}, where coverage is key and
% precision secondary; e.g.\ for consumer-grade machine translation like \textsf{Google translate}.
% 
% In this paper, we address tools that help attack the coverage problem in symbolic/semantic
% approaches in practice. The producer task we use as a case study is that of adaptive
% learning assistants for STEM subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
% \ALeA uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
% variant of {\LaTeX} that allows to embed semantic annotations -- for knowledge
% representation, and generates learner-adaptive learning objects instrumented with
% learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
% \ednote{MK@MK: continue to describe the size and distribution, and
%   authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
%   \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
%   relative sizes.}
% 
% We contend that while education -- especially higher education -- as a whole is a
% wide-coverage task for society education of individual learners in specific domains --
% especially, if we want to tailor the educational offerings to said individual/domain -- is
% a producer task, at least until ML-based methods reach the precision and explainability
% required by the ethics of teaching.

\section{Semantic Authoring in a Content Commons}\label{sec:semauth}

@@

\subsection{Running Example: Annotating  ``Terminal''}
With \sTeX, authors can add semantic references to technical terms,
which enables straight-forward learning support services like
showing the definition of a term, but also informs the system
about the content/prerequisites of a learning object.
Adding these references is a key practical problem when converting presentational {\LaTeX}
course materials into semantic documents.

@@

Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
the term ``terminal'' with the (semantic) concept of ``a goal state of a search
problem''. The latter can either be defined earlier in the course or in the domain model
(or both).

% The main practical problem of annotating presentational {\LaTeX} course materials into
% semantic documents that carry enough information to support meaningful learning support
% services is to add semantic references to technical terms from the domain jargon.

@@

To annotate the string ``terminal'' in the learning object, the author has to be aware
that it is a technical term, that the module from \cref{fig:state-space} exists, know the
symbol name and its URL\ednote{path/URI/identifier?}, and manage redundancy of the imports -- i.e. only
adding the \lstinline|\importmodule| directive if it is not already (recursively) implied
and possibly removing directives that become redundant by the new one.  At ca 150,000
words in a semester's worth of lecture notes and ca. 10--15\%\ednote{can we actually
  measure this for AI?} of technical terms a rather daunting task.

\sTeX authoring is already supported by a VSCode IDE plugin \cite{sTeX-IDE:git} that
analyzes annotations on the fly, displays the underlying reference semantics, reports
errors and redundancies, and even offers a concept search interface.
But these features only support already annotated terms, so they do not solve the annotation problem above.\ednote{
    last paragraph doesn't fit here anymore, as it only motivates \snify
}

\subsection{Semantic Authoring as a Distinct Problem}
We contend that semantic authoring is a problem that is distinct from traditional
(informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
while sharing some of the aspects of either.

@@

\paragraph{Semantic Authoring}
In semantic authoring, e.g. the \sTeX sources of educational content for the \ALeA system
presented above, the base of the text is classical narrative text to which most of the
constraints of classical authoring apply.
The exception is the tailoring to the competencies of the assumed reader
that traditionally happens during authoring, but which can (partially) be automated by systems like
\ALeA. Note that it is not a restriction that \ALeA is initially targeted to tertiary
education; the underlying mechanisms generalize to all scientific communication and
technical documents. After all, the purpose is to transfer information (instruct), even if
the roles of instructor and instructee are not predetermined by academic seniority.

@@

\paragraph{Semantic Authoring in \sTeX}
How this can concretely work depends on the particular semantic framework, the specific
annotations we want to support, and the semantics embedded in the content commons. In the
\sTeX framework, where we semantically annotate formulae and technical terms with pointers
to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
in sentences and glyphs in formulae with the verbalizations and notations in the content
commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
formulae via notations has been presented by Vre{\v{c}}ar et al. in
\cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
annotation of technical terms via verbalizations.\ednote{move this in the next section}

\section{Existing and Envisioned Authoring Support for \sTeX/\ALeA}\label{sec:tools}

\section{The \snify System}\label{sec:snify}

@@

\section{Practical Evaluation}\label{sec:eval}

@@

\section{Conclusion}\label{sec:conclusion}
